---
title: "maxentcpp Performance Benchmark Report"
author: "maxentcppCompTest"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Overview

This report benchmarks the training and prediction performance of
**maxentcpp** (C++17 reimplementation) against **dismo** (Java Maxent
via rJava) on the bundled mock fixtures.

### Environment

```{r env-info}
cat("R version:", R.version.string, "\n")
cat("Platform:", Sys.info()["sysname"], Sys.info()["machine"], "\n")
cat("maxentcpp:", as.character(packageVersion("maxentcpp")), "\n")
if (requireNamespace("dismo", quietly = TRUE))
    cat("dismo:", as.character(packageVersion("dismo")), "\n")
cat("Date:", format(Sys.time(), "%Y-%m-%d %H:%M %Z"), "\n")
```

## Data Preparation

```{r data-prep}
library(maxentcppCompTest)
library(maxentcpp)
library(microbenchmark)

paths <- mock_raster_paths()
occ   <- mock_occurrences()

# maxentcpp data
g1 <- maxent_read_asc(paths["bio1"])
g2 <- maxent_read_asc(paths["bio2"])
m1 <- maxent_grid_to_matrix(g1)
m2 <- maxent_grid_to_matrix(g2)
info <- maxent_grid_info(g1)

bio1_vec <- as.numeric(t(m1))
bio2_vec <- as.numeric(t(m2))
n <- length(bio1_vec)

occ_indices <- vapply(seq_len(nrow(occ)), function(i) {
    row_i <- floor((info$yll + info$nrows * info$cellsize - occ$lat[i]) /
                       info$cellsize)
    col_i <- floor((occ$lon[i] - info$xll) / info$cellsize)
    row_i <- max(0L, min(as.integer(row_i), info$nrows - 1L))
    col_i <- max(0L, min(as.integer(col_i), info$ncols - 1L))
    as.integer(row_i * info$ncols + col_i)
}, integer(1L))

cat(sprintf("Dataset: %d cells, %d variables, %d samples\n",
            n, 2L, nrow(occ)))
```

```{r check-dismo}
has_dismo <- requireNamespace("dismo", quietly = TRUE) &&
    requireNamespace("rJava", quietly = TRUE) &&
    requireNamespace("raster", quietly = TRUE)

if (has_dismo) {
    jar_path <- file.path(system.file(package = "dismo"), "java", "maxent.jar")
    has_dismo <- file.exists(jar_path)
}
cat("dismo + Java available:", has_dismo, "\n")
```

## Training Benchmark

### Symmetric Fixture (100 cells)

```{r train-bench}
bench_cpp_train <- function() {
    f1 <- maxent_linear_feature(bio1_vec, "bio1")
    f2 <- maxent_linear_feature(bio2_vec, "bio2")
    fs <- maxent_featured_space(n, occ_indices, list(f1, f2))
    maxent_fit(fs, max_iter = 500L, convergence = 1e-5,
               beta_multiplier = 1.0)
}

n_reps <- 20L

if (has_dismo) {
    env_raster <- raster::stack(paths["bio1"], paths["bio2"])
    occ_coords <- occ[, c("lon", "lat")]

    # Warm up JVM
    suppressMessages(try(
        dismo::maxent(env_raster, occ_coords, silent = TRUE,
                      args = c("nowarnings", "notooltips",
                               "maximumiterations=10")),
        silent = TRUE))

    bench_dismo_train <- function() {
        dismo::maxent(
            env_raster, occ_coords, silent = TRUE,
            args = c("nowarnings", "notooltips",
                     "linear=true", "quadratic=false",
                     "hinge=false", "threshold=false",
                     "product=false",
                     "maximumiterations=500",
                     "convergencethreshold=1e-5",
                     "betamultiplier=1.0"))
    }

    train_results <- microbenchmark(
        maxentcpp = bench_cpp_train(),
        dismo     = bench_dismo_train(),
        times = n_reps, setup = gc(verbose = FALSE))
} else {
    train_results <- microbenchmark(
        maxentcpp = bench_cpp_train(),
        times = n_reps, setup = gc(verbose = FALSE))
}

print(train_results)
```

```{r train-summary}
train_summary <- summary(train_results)
train_summary$median_ms <- train_summary$median / 1e6
train_summary$mean_ms   <- train_summary$mean / 1e6
knitr::kable(train_summary[, c("expr", "median_ms", "mean_ms", "neval")],
             digits = 2, caption = "Training time (ms)")
```

```{r train-plot, fig.width=7, fig.height=4}
boxplot(train_results, unit = "ms", log = FALSE,
        main = "Training Time Distribution",
        ylab = "Time (ms)")
```

### Asymmetric Fixture (non-trivial lambdas)

```{r train-asym}
paths_asym <- mock_raster_paths_asym()
occ_asym   <- mock_occurrences_asym()

g1a <- maxent_read_asc(paths_asym["bio1"])
g2a <- maxent_read_asc(paths_asym["bio2"])
m1a <- maxent_grid_to_matrix(g1a)
m2a <- maxent_grid_to_matrix(g2a)

bio1_asym <- as.numeric(t(m1a))
bio2_asym <- as.numeric(t(m2a))

occ_indices_asym <- vapply(seq_len(nrow(occ_asym)), function(i) {
    row_i <- floor((info$yll + info$nrows * info$cellsize -
                        occ_asym$lat[i]) / info$cellsize)
    col_i <- floor((occ_asym$lon[i] - info$xll) / info$cellsize)
    row_i <- max(0L, min(as.integer(row_i), info$nrows - 1L))
    col_i <- max(0L, min(as.integer(col_i), info$ncols - 1L))
    as.integer(row_i * info$ncols + col_i)
}, integer(1L))

bench_cpp_train_asym <- function() {
    f1 <- maxent_linear_feature(bio1_asym, "bio1")
    f2 <- maxent_linear_feature(bio2_asym, "bio2")
    fs <- maxent_featured_space(n, occ_indices_asym, list(f1, f2))
    maxent_fit(fs, max_iter = 500L, convergence = 1e-5,
               beta_multiplier = 1.0)
}

asym_results <- microbenchmark(
    maxentcpp_asym = bench_cpp_train_asym(),
    times = n_reps, setup = gc(verbose = FALSE))

print(asym_results)
```

## Prediction Benchmark

```{r predict-bench}
# Train maxentcpp model for prediction
f1 <- maxent_linear_feature(bio1_vec, "bio1")
f2 <- maxent_linear_feature(bio2_vec, "bio2")
fs <- maxent_featured_space(n, occ_indices, list(f1, f2))
maxent_fit(fs, max_iter = 500L, convergence = 1e-5, beta_multiplier = 1.0)

# Build feature matrix
f1p <- maxent_linear_feature(bio1_vec, "bio1")
f2p <- maxent_linear_feature(bio2_vec, "bio2")
feat_mat <- cbind(
    vapply(seq_len(n), function(i) maxent_feature_eval(f1p, i), numeric(1)),
    vapply(seq_len(n), function(i) maxent_feature_eval(f2p, i), numeric(1)))

bench_cpp_predict <- function() {
    maxent_predict_model(fs, feat_mat)
}

if (has_dismo) {
    dismo_model <- dismo::maxent(
        env_raster, occ_coords, silent = TRUE,
        args = c("nowarnings", "notooltips",
                 "linear=true", "quadratic=false",
                 "hinge=false", "threshold=false",
                 "product=false",
                 "maximumiterations=500",
                 "convergencethreshold=1e-5",
                 "betamultiplier=1.0"))

    bench_dismo_predict <- function() {
        dismo::predict(dismo_model, env_raster)
    }

    pred_results <- microbenchmark(
        maxentcpp = bench_cpp_predict(),
        dismo     = bench_dismo_predict(),
        times = n_reps, setup = gc(verbose = FALSE))
} else {
    pred_results <- microbenchmark(
        maxentcpp = bench_cpp_predict(),
        times = n_reps, setup = gc(verbose = FALSE))
}

print(pred_results)
```

```{r predict-summary}
pred_summary <- summary(pred_results)
pred_summary$median_ms <- pred_summary$median / 1e6
pred_summary$mean_ms   <- pred_summary$mean / 1e6
knitr::kable(pred_summary[, c("expr", "median_ms", "mean_ms", "neval")],
             digits = 2, caption = "Prediction time (ms)")
```

```{r predict-plot, fig.width=7, fig.height=4}
boxplot(pred_results, unit = "ms", log = FALSE,
        main = "Prediction Time Distribution",
        ylab = "Time (ms)")
```

## Numerical Verification

```{r numerical-verify}
cpp_preds <- maxent_predict_model(fs, feat_mat)
cat(sprintf("maxentcpp: min=%.6f, max=%.6f, mean=%.6f\n",
            min(cpp_preds), max(cpp_preds), mean(cpp_preds)))

if (has_dismo) {
    dismo_preds <- as.numeric(raster::values(
        dismo::predict(dismo_model, env_raster)))
    cat(sprintf("dismo:     min=%.6f, max=%.6f, mean=%.6f\n",
                min(dismo_preds, na.rm = TRUE),
                max(dismo_preds, na.rm = TRUE),
                mean(dismo_preds, na.rm = TRUE)))

    cmp <- compare_maxent_predictions(cpp_preds, dismo_preds)
    knitr::kable(data.frame(
        Metric = c("Spearman rho", "Pearson r",
                    "Max rank diff", "Agreement"),
        Value  = c(sprintf("%.6f", cmp$spearman_cor),
                   sprintf("%.6f", cmp$pearson_cor),
                   sprintf("%.6f", cmp$max_abs_diff_rank),
                   ifelse(cmp$agreement, "YES", "NO"))
    ), caption = "Cross-implementation agreement")
}
```

## Combined Results Table

```{r combined-table}
rows <- list()

# Training
ts <- summary(train_results)
ts$median_ms <- ts$median / 1e6
for (i in seq_len(nrow(ts))) {
    rows[[length(rows) + 1]] <- data.frame(
        Dataset = "Mock (100 cells)",
        Phase   = "Train",
        Impl    = as.character(ts$expr[i]),
        Median_ms = ts$median_ms[i],
        stringsAsFactors = FALSE)
}

# Prediction
ps <- summary(pred_results)
ps$median_ms <- ps$median / 1e6
for (i in seq_len(nrow(ps))) {
    rows[[length(rows) + 1]] <- data.frame(
        Dataset = "Mock (100 cells)",
        Phase   = "Predict",
        Impl    = as.character(ps$expr[i]),
        Median_ms = ps$median_ms[i],
        stringsAsFactors = FALSE)
}

combined <- do.call(rbind, rows)
knitr::kable(combined, digits = 2,
             caption = "Combined benchmark results (median, ms)")
```

## Session Info

```{r session-info}
sessionInfo()
```
